Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-07 extended_catalog/paper/extended_catalog.tex
markus changes

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope (Fermi)}. Measuring the extension
of a LAT source is important for uniquely identifying it.  We provide
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We present on the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi LAT Catalog (2FGL).
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope (Fermi). It has been surveying the
$\gamma$-ray sky since June 2008.  Fermi is well suited as a survey
instrument because of its broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\approx 2.4 \sr$), and large effective area
($\approx 8000 \cm^2$ at $>1 \gev$).

Using two years of all sky surveying data, the LAT Collaboration
published a catalog of 1873 \gev sources\cite{second_cat}. Many of these can
be positionally associated with a variety of source classes including
Active Galactic Nuclei (AGN), pulsars, galaxies, and supernova
remnants (SNRs).  While many of these classes can be spatially
resolved when observed at other frequencies, detection an extension at
\gev energies by the LAT is complicated by the rather large
point-spread function (PSF) of the instrument (the angular resolution
for events converting in the front part of the detector, defined as
the 68\% containment radius for a point source, is approximately
$3.5^{\circ}$ at 100 \mev, improving to about 0.1$^{\circ}$ at 10 \gev).

Several of these source classes are extended (and close) enough to
potentially be extended. In particular supernova remnants, pulsar wind
nebulae, molecular clouds, normal galaxies, galaxy clusters or dark
matter satellites can (in favorable cases) be expected to be extended
beyond the LAT's PSF. The LAT collaboration has previously reported on
the discovered of five spatially extended supernova remnants, IC443,
W28, W44, W51C, and RX J1713.7-3946
(\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended
pulsar wind nebulae (PWNe) were detected: MSH 15-52, Vela~X, and
HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well as
two close-by galaxies, the Large Magellanic and Small Magalenic Cloud
(\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

In addition to these \gev $\gamma$-ray detections, a variety of
spatially extended sources have been detected in the Galactic plane at
\tev energies using Air Cherenkov detectors. Most prominent was a
survey of the Galactic plane using the High Energy Stereoscopic System
(H.E.S.S) which presented 14 spatially extended sources
(\cite{hess_plane_survey}) sources with extensions varying from
$~0.1\deg$ to $~0.25\deg$. In fact, within our Galaxy only very few
sources (most notably the $\gamma$-ray binaries
LS\,5039 (\cite{HESSLS5039}), LS I+61-303 (\cite{MAGICLSI, VERITASLSI}) and
HESS\,J0632+057 (\cite{HESS0632}) are amongst the few Galactic sources
for which no significant extension was detected at \tev
energies. Given these findings, it can be expected that if the source
population at \gev energies has similar properties, that at least some
of the Fermi-LAT sources in our Galaxy should be spatially extended at
\gev energies.

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the Fermi-LAT. Finding a matching morphology can be used
to uniquely identify the source.  Because of the strongly varying PSF
of the LAT, the spatial and spectral information about a source do not
simply decouple. Therefore, a biased spatial model will bias the
spectral model of the source and studying the extension of the source
can correct for this.  Furthermore, correctly modeling a source's
extension is important for improving the model of the sky and
removing excess residuals.  These residuals manifested themselves in
1FGL as spurious point sources in the region region surrounding the
Large Magellanic Cloud (\cite{first_cat}).  For this reason, we are
motivated to take the 1873 sources that were detected in 2FGL and
apply a search for those sources which are spatially extended.

@@

Morphological studies of sources in the \gev energy range
using the LAT are challenging because of the significantly 
energy-dependent PSF and because of systematic errors in the Galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are typically limited by statistics. 
A detailed analysis procedure must be used to maximize our
sensitivity over all energies.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is maximized by fitting
parameters of the model.  The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position
and extension can be simultaneously fit.

>>>>

\begin{abstract}
We present a new method for analyzing 
the spatial extension of sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em Fermi Gamma-ray Space Telescope (Fermi)}. 
The source extension is an important input to correctly 
associate LAT sources with their counterparts at other
wavelengths. We provide
a series of Monte Carlo studies to validate this tool and calculate
the LAT's detection threshold to spatially extended sources.  We then
apply this tool to test all sources in the two year source catalog for
extension. We present on the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second Fermi-LAT Catalog (2FGL).
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the Fermi
Gamma-Ray Space Telescope (FGST). It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi has a broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\approx 2.4 \sr$), and large effective area
($\approx 8000 \cm^2$ at $>1 \gev$).

Using two years of all sky surveying data, the LAT Collaboration
published a catalog of 1873 \gev sources\cite{second_cat}. Many of these can
be positionally associated with a variety of source classes including
Active Galactic Nuclei (AGN), pulsars, galaxies, and supernova
remnants (SNRs).  While many of the LAT sources can be
spatially
resolved when observed at other frequencies, detection of 
an extension at
\gev energies by the LAT is complicated by the rather large
point-spread function (PSF) of the instrument. The angular resolution
for events converting in the front part of the detector, defined as
the 68\% containment radius for a point source, is approximately
$3.5^{\circ}$ at 100 \mev, improving to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident
with LAT sources exhibit extensiosn at other
wavelengths which are larger or comparable to the LAT PSF.
In particular certain SNRs, pulsar wind
nebulae (PWNe), molecular clouds, normal galaxies, 
galaxy clusters or dark
matter satellites can therefore be expected
to be observed as spatially extended sources in the LAT.
The LAT collaboration has previously reported on
the discovered of five spatially extended supernova remnants, IC443,
W28, W44, W51C, and RX J1713.7-3946
(\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended
PWNe were detected: MSH 15-52, Vela~X, and
HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well as
two close-by galaxies, the Large Magellanic and Small Magalenic Cloud
(\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

For each of these sources, the extension was determined based on
dedicated analysis of the particular source or region of interest
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented here, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using Air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which presented 14 spatially extended sources
 with extensions varying from
$~0.1\deg$ to $~0.25\deg$ (\cite{hess_plane_survey}). In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 (\cite{HESSLS5039}), LS I+61-303 (\cite{MAGICLSI, VERITASLSI})
and HESS\,J0632+057 (\cite{HESS0632}) are amongst the few Galactic sources
for which no significant extension was detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the Fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the identification of LAT sources.  Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Because of the
strongly varying PSF of the LAT, the spatial and spectral information
about a source do not simply decouple. A biased spatial model will bias
the spectral model of the source. In particular, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud (\cite{first_cat}).
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the significantly energy-dependent PSF.
Additional complications arise close to the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at lower energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately 5.1\deg
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is 0.14\deg at 10 \gev
(\cite{on_orbit_calibration}).  The high resolution of the higher
energy photons in the LAT energy range however is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(see below).

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is maximized,
given a parametrized spatial and spectral model of 
the source and its surrounding region.
The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position,
extension, and spectrum can be simultaneously fit.
